### 2018-03-20 doc/naaclhlt2018-latex/paper.tex
add paper file and start feature description

%
% File naaclhlt2018.tex
%
%% Based on the style files for NAACL-HLT 2018, which were
%% Based on the style files for ACL-2015, with some improvements
%%  taken from the NAACL-2016 style
%% Based on the style files for ACL-2014, which were, in turn,
%% based on ACL-2013, ACL-2012, ACL-2011, ACL-2010, ACL-IJCNLP-2009,
%% EACL-2009, IJCNLP-2008...
%% Based on the style files for EACL 2006 by 
%%e.agirre@ehu.es or Sergi.Balari@uab.es
%% and that of ACL 08 by Joakim Nivre and Noah Smith

\documentclass[11pt,a4paper]{article}
\usepackage[hyperref]{naaclhlt2018}
\usepackage{times}
\usepackage{latexsym}
\usepackage{soul}

\usepackage{url}

\aclfinalcopy % Uncomment this line for the final submission
%\def\aclpaperid{***} %  Enter the acl Paper ID here

%\setlength\titlebox{5cm}
% You can expand the titlebox if you need extra space
% to show all the authors. Please do not make the titlebox
% smaller than 5cm (the original size); we will check this
% in the camera-ready version and ask you to change it back.

\newcommand\BibTeX{B{\sc ib}\TeX}

\title{Title: Something snappy about the SLAM competition}

\author{Alexander S. Rich\qquad Pamela Osborn Popp\qquad David J. Halpern\\
  \textbf{Anselm Rothe\qquad Todd M. Gureckis} \\
  Department of Psychology, New York University \\
  {\tt \{asr443,pamop,david.halpern,anselm,todd.gureckis\}@nyu.edu} \\}

\date{}

\begin{document}
\maketitle
\begin{abstract}
\end{abstract}

\section{Introduction}

Blah Blah \cite{slam18}

\section{Task Approach}

We approached the task as a binary classification problem over instances (i.e.,
single words within an exercise). Our solution can be divided into two
components---constructing a set of features that is highly informative about
whether the user will answer an instance correctly, and designing a model that
can achieve high performance using this feature set.

\subsection{Feature Engineering}

We used a variety of features, including features directly present in the
training data, features constructed using the training data, and features that
use information external to the training data. Except where otherwise specified,
categorical variables were one-hot encoded.

\subsubsection{Exercise features}

We encoded the exercise client, session, format, and time (i.e., number of
seconds to complete the exercise), as well as the exercise number and number of
days since start of usage.

\subsubsection{Word features}

Using spaCy\footnote{https://spacy.io/}, we lemmatized each word to produce a root word. Both the root
word token and the original token were used as categorical features. Due to
their high cardinality, these features were not one-hot encoded but were
preserved in single columns and handled in this form by the model (as described
below).

Along with the tokens themselves we encoded instance word's part of speech,
morphological features, and dependency edge label. (We noticed that some words
in the original dataset were paired with the wrong morphological features,
particularly near where punctuation had been removed from the sentence. To fix,
this, we re-processed the data using Google SyntaxNet\footnote{https://github.com/ljm625/syntaxnet-rest-api}.)

We also encoded word length and several word characteristics gleaned from
external data sources. \hl{PAM AND DAVID FILL IN HERE WITH FREQ, LEVENSHTEIN, AND
AOA}.

\subsubsection{User features}

Just as we did for word tokens, we encoded the user ID as a single-column,
high-cardinality feature. We also calculated several other user-level features.
\hl{TODD AND ANSELM HERE}

\subsubsection{Positional features}

To account for the effects of surrounding words on the difficulty of an
instance, we created several features related to the instance word's context in
the exercise. These included the token of the previous word, the next word, and
the instance word's root in the parse tree, all stored in single columns as with
the instance token itself. We also included the part of speech of each of these
context words as additional features. When there was no previous word, next word, or parse
root word, a special {\tt None} token or {\tt None} part of speech was used.

\subsubsection{Temporal features}

A user's probability of succeeding on an instance is likely related to their
prior experience with that instance. To capture this, we calculated several
features related to past experience. We encoded the number of times the
current exercise's exact sentence had been seen before by the user. \hl{is this
  right Todd?} We also encoded a set of features recording past experience with
the particular instance word. These features were encoded separately for the
instance token and for the instance root word created by lemmatization.

For each token (and root) we tracked user performance through four weighted
error averages. At the user's first encounter of the token, each error term $E$ starts at
zero. After an encounter with an instance of the token with label $L$, it is
updated according to the equation

\[
E \leftarrow E + \alpha (L - E)
\]

where $\alpha$ determines the speed of error updating. The four feature weighted
error terms use $\alpha = \{.3, .1, .03, .01\}$, allowing both short-run and
long-run changes in a user's error rate with a token to be tracked. Note that in
cases where a token appears multiple times in an exercise, a single update of
the error features is conducted using the mean of the token labels.
Along with the error tracking features, for each token we calculated the number
of labeled, unlabeled, and total encounters; time since last labeled encounter and
last encounter; and whether the instance is the first encounter with the
token.

In the training data, all instances are labeled as correct or incorrect, so the
label for the previous encounter is always available. In the test data, labels
are unavailable, so predictions must be made using a mix of labeled and
unlabeled past encounters. To generate training-set features that are 
comparable to test-set features, we selectively ignored some labels when encoding temporal features on
the training set. Specifically, for each user we first calculated the number of
exercises $n$ in the true test set. Then, when encoding the features for each
instance, we selected a random integer $r$ in the range $[1,n]$, and ignored labels
in the prior $r$ exercises. That is, we encode features for the current instance
as though other instances in those prior exercises were unlabeled, and ignore
updates to the error averages from those exercises. The result of this process
is that each instance in the training set is encoded as though it were between
one and $n$ exercises into the test set.

\subsection{Modeling}

\section{Feature Removal Experiments}

\bibliography{paper}
\bibliographystyle{acl_natbib}

\end{document}

### 2018-03-21 doc/naaclhlt2018-latex/paper.tex
add modeling section to paper

@@
 
+After featurizing the training data, we trained gradient boosting decision
+tree (GBDT) models to minimize log loss. GBDT works by
+iteratively building regression trees, each of which seeks to minimize the
+residual loss from prior trees. This allows it to capture non-linear effects
+and high-order interactions among features. We used the LightGBM\footnote{http://lightgbm.readthedocs.io/} implementation
+of GBDT \cite{ke2017lightgbm}.
+
+For continuous-valued features, GBDT can split a leaf at any point, creating
+different predicted values above and below that threshold. For categories that
+are one-hot encoded, it can split a leaf on any of the category's features. This
+means that for a category with thousands of values, potentially thousands of
+tree splits would be needed to capture its relation to the target. Fortunately,
+LightGBM implements an algorithm for partitioning the values of a categorical
+feature into two groups based on their relevence to the current loss, and create
+a single split to divide those groups \cite{fisher1958grouping}. Thus, as
+alluded to above, high-cardinality features like token and user were encoded as
+single columns and handled as categories by LightGBM.
+
+We trained a model for each of the three language tracks of {\tt en\_es}, {\tt es\_en},
+and {\tt fr\_en}, and also trained a model on the combined data from
+all three tracks, adding an additional ``language'' feature. Following model
+training, we averaged the predictions of each single-language model with that of
+the all-language model to form our final predictions.
+
+To tune model hyper-parameters and evaluate the usefulness of features, we first
+trained the models on the ``train'' data set and evaluated them on the ``dev''
+data set. Once the model structure was finalized, we trained on the combined
+``train'' and ``dev'' data and produced predictions for the ``test'' data. The
+LightGBM hyperparameters used for each model are listed in Table~\ref{lightgbm-params}.
+
+\begin{table}[t!]
+\begin{center}
+\begin{tabular}{|l|llll|}
+  % \hline   & \multicolumn{3}{l}{\bf Value} & \\
+  \hline \bf Parameter & {\tt fr\_en} & {\tt en\_es} & {\tt es\_en} & {\tt all} \\ \hline
+  num\_leaves & 256 & 512 & 512 & 1024 \\
+  learning\_rate & .05 & .05 & .05 & .05 \\
+  min\_data\_in\_leaf & 100 & 100 & 100 & 100 \\
+  num\_boost\_rounds & 750 & 650 & 600 & 750 \\
+  cat\_smooth & 200 & 200 & 200& 200 \\
+  feature\_fraction & .7 & .7 & .7 & .7 \\
+  max\_cat\_threshold & 32 & 32 & 32& 64 \\
+
+\hline
+\end{tabular}
+\end{center}
+\caption{\label{lightgbm-params} Parameters of final LightGBM models. See
+  LightGBM documentation for more information; all other parameters were left at
+their default values.}
+\end{table}
+
+\subsection{Performance}
+
+The AUROC of our final predictions was $.8585$ on {\tt en\_es}, $.8350$ on {\tt
+  es\_en}, and $.8540$ on {\tt fr\_en}. We did not attempt to optimize the model's
+F1 score, but the model's F1 score could likely be improved (at the cost of
+increased log loss) by finding the
+rescaling of the ``dev'' predicted probabilities that maximized the F1 score at
+the 0.5 threshold, and applying this rescaling to the ``test'' predicted probabilities.
+
 \section{Feature Removal Experiments}

### 2018-03-21 doc/naaclhlt2018-latex/paper.tex
word freq and edit distance description

@@
 
-Along with the tokens themselves we encoded instance word's part of speech,
+Along with the tokens themselves we encoded \hl{each} instance word's part of speech,
 morphological features, and dependency edge label. (We noticed that some words
 in the original dataset were paired with the wrong morphological features,
-particularly near where punctuation had been removed from the sentence. To fix,
+particularly near where punctuation had been removed from the sentence. To fix
 this, we re-processed the data using Google SyntaxNet\footnote{https://github.com/ljm625/syntaxnet-rest-api}.)
@@
 We also encoded word length and several word characteristics gleaned from
-external data sources. \hl{PAM AND DAVID FILL IN HERE WITH FREQ, LEVENSHTEIN, AND
-AOA}.
+external data sources.
+The word frequency effect suggests that uncommon words are harder to process than common words; readers will look longer at low-frequency words and perform worse in word-identification tasks for these than for high-frequency words \cite{rayner1998eye}.
+We therefore included a feature that encoded the frequency of each word in the language being acquired, calculated from \citet{robert_speer_2017_998161}.
+Additionally, cognates, or words sharing a common linguistic derivation, are easier to learn than words with dissimilar translations \cite{de2000hard}.
+As an approximate measure of linguistic similarity, we used the Levenshtein or edit distance between the word tokens and their translations scaled by the length of the longer word.
+We found translations using Google Translate\footnote{https://cloud.google.com/translate/} and calculated the Levenshtein distance to reflect the letter-by-letter similarity of the word and its translation \cite{hyyro2001explaining}.
+\hl{DAVID FILL IN HERE WITH AOA}.
 

### 2018-03-22 doc/naaclhlt2018-latex/paper.tex
some words…

@@
 
-\title{Title: Something snappy about the SLAM competition}
+\title{Modeling Second-Language Learning from a Psychological Perspective}
 
@@
 \begin{abstract}
+This is the abstract
 \end{abstract}
@@
 
-Blah Blah \cite{slam18}
+Educational software that aims to teach people new skills, languages, and
+even academic subjects have become increasingly popular.  The wide-spread
+deployment of these tools has created interesting opportunities to study
+the process of learning in extremely large samples of learners in naturalistic
+situations. The Duolingo shared task on Second Lanugage Acquisition Modeling (SLAM)
+was a competitive modeling challenge run in early 2018~\cite{slam18}.
+The challenge, organized by Duolingo\footnote{http://duolingo.com}, a popular second
+language learning app, was
+to use log data from thousand of users completing millions of exercises 
+predict patterns of future translation mistakes in held-out data.  The data set was
+divided into three sets covering students learning to translate English to Spanish (ES$\to$EN),
+Spanish to English (ES$\to$EN), and French to English (FR$\to$EN).
+This paper reports the approach used by our team which ended in third place 
+for the ES$\to$EN data set, second place for the ES$\to$EN data set, and third place
+for FR$\to$EN.
 
+Our lab's interest in this competition comes from our emphasis on studying human learning
+and memory from a psychological perspective.  While the Duolingo SLAM dataset 
+heavily emphasizes language learning (e.g., syntax, grammar, semantics), it also touches on
+basic learning and memory phenomena (e.g., retrieval of the translated
+word from memory).  The study of the psychological processes supporting second-language 
+acquisition has long been a topic of interest in the learning and memory literature~\cite[e.g.,]{Atkinson:1972rm,Atkinson:1972rz,Pavlik:2008rm}.
+In fact, much of the classic work on psychology of memory involved aspects of verbal
+language learning, which helped to identify a number of important and robust phenomena
+in memory including the retention curve and the advantage for spaced
+over massed practice~\cite{Ruth:1928aa,Rubin:1996aa,Cepeda:2006aa}.
+
+\hl{ Todd: I want to transition to something about how our approach tried to stick to some
+psychological principles but i'm not sure... maybe need to wait until al the later sections
+are fleshed out to lay out what the "concept" is behind our approach}.
+
+\hl{One idea is to talk about while the research on learnign and memory is important
+it might be hard to graft those principles into datasets like these because of 
+subject variability, etc... As a result it is unlikely that straightfoward
+cognitive models will work but instead need to combine maching learning methods
+that can create much more complex rules and decisions with features that reflect
+aspects of what makes something easy or hard to learn.}
+
+\cite{Tubridy:2018aa}
+\cite{Tubridy:2018ab}
+\cite{Corbett1995bkt}
+\cite{Roediger2006te,Anderson1994rif,Mozer:2009cs}
 \section{Task Approach}
@@
 features related to past experience. We encoded the number of times the
-current exercise's exact sentence had been seen before by the user. \hl{is this
-  right Todd?} We also encoded a set of features recording past experience with
+current exercise's exact sentence had been seen before by the user. 
+We also encoded a set of features recording past experience with
 the particular instance word. These features were encoded separately for the
 instance token and for the instance root word created by lemmatization.
+\cite{Perruchet:1998nw,Chun:1999gt} \hl{chun is good, maybe something other
+that PARSER as it is more a statistcal learning model}
 
@@
 
+\section{Acknowledgments}
+This research was supported by NSF grant DRL-1631436 and
+seed funds from the NYU Dean for Science.
+
 \bibliography{paper}

### 2018-03-23 doc/naaclhlt2018-latex/paper.tex
small edits

@@
 
-We encoded the exercise client, session, format, and time (i.e., number of
-seconds to complete the exercise), as well as the exercise number and number of
-days since start of usage.
+We encoded the exercise number, client, session, format, and duration (i.e., number of
+seconds to complete the exercise), as well as the time since the user started using Duolingo for the first time.
 
@@
 Additionally, cognates, or words sharing a common linguistic derivation, are easier to learn than words with dissimilar translations \cite{de2000hard}.
-As an approximate measure of linguistic similarity, we used the Levenshtein or edit distance between the word tokens and their translations scaled by the length of the longer word.
+As an approximate measure of linguistic similarity, we used the Levenshtein edit distance between the word tokens and their translations scaled by the length of the longer word.
 We found translations using Google Translate\footnote{https://cloud.google.com/translate/} and calculated the Levenshtein distance to reflect the letter-by-letter similarity of the word and its translation \cite{hyyro2001explaining}.
@@
 high-cardinality feature. We also calculated several other user-level features.
-\hl{TODD AND ANSELM HERE}
+
+\hl{ANSELM HERE}
+
+Feature 1: Motivation %compute_motivation
+
+Feature 2: Consistency %compute_usage_entropy
 
@@
 \begin{table}[t!]
+\small
 \begin{center}

### 2018-03-24 preprocess_to_pickle.py
add code to run create pickled data and run lesion experiments

@@ -1,0 +1,72 @@
+from processing import build_data
+import pickle
+
+
+data = build_data(
+    'all',
+    [
+        'data/data_{0}/{0}.slam.20171218.train.new'.format('en_es'),
+        'data/data_{0}/{0}.slam.20171218.train.new'.format('fr_en'),
+        'data/data_{0}/{0}.slam.20171218.train.new'.format('es_en')
+    ],
+    [
+        'data/data_{0}/{0}.slam.20171218.dev.new'.format('en_es'),
+        'data/data_{0}/{0}.slam.20171218.dev.new'.format('fr_en'),
+        'data/data_{0}/{0}.slam.20171218.dev.new'.format('es_en')
+    ],
+    labelfiles=[
+        'data/data_{0}/{0}.slam.20171218.dev.key'.format('en_es'),
+        'data/data_{0}/{0}.slam.20171218.dev.key'.format('fr_en'),
+        'data/data_{0}/{0}.slam.20171218.dev.key'.format('es_en')
+    ])
+train_x, train_ids, train_y, test_x, test_ids, test_y = data
+
+
+word_feat = 'token'
+word_stats = {}
+langlist = ['en_es', 'fr_en', 'es_en']
+for l in langlist:
+    with open('data/'+l+'_wordwordfeats.txt', 'r') as f:
+        for line in f.readlines():
+            line = line.split(',')
+            # add language identifier tag to end of word,
+            # as is done in features
+            word_stats[line[0].lower()+'_'+l[:2]] = {
+                'frequency': float(line[2]),
+                'levenshtein': int(line[3]),
+                'leven_frac': float(line[4]),
+                'aoa': float(line[5])
+            }
+
+for d in train_x + test_x:
+    word = d[word_feat].lower()
+    if word in word_stats:
+        stats = word_stats[word]
+        d['frequency'] = stats['frequency']
+        d['levenshtein'] = stats['levenshtein']
+        d['leven_frac'] = stats['leven_frac']
+        d['aoa'] = stats['aoa']
+
+cat_features = ['token', 'root', 'user',
+                'prev_token', 'next_token', 'parseroot_token']
+for key in cat_features:
+    val_dict = {}
+    val_idx = 0
+    for d in train_x + test_x:
+        t = d[key]
+        if t in val_dict:
+            d[key] = val_dict[t]
+        else:
+            val_dict[t] = val_idx
[…]

### 2018-03-24 run_lesion.py

@@ -1,0 +1,96 @@
+import os.path
+import sys
+import pickle
+from sklearn.feature_extraction import DictVectorizer
+from sklearn.metrics import roc_auc_score
+import lightgbm as lgb
+import argparse
+
+parser = argparse.ArgumentParser()
+parser.add_argument('--lesion', default='none')
+args = vars(parser.parse_args())
+lesion_type = args['lesion']
+
+max_bytes = 2**31 - 1
+
+# read
+bytes_in = bytearray(0)
+input_size = os.path.getsize('alldata.p')
+with open('alldata.p', 'rb') as f_in:
+    for _ in range(0, input_size, max_bytes):
+        bytes_in += f_in.read(max_bytes)
+data = pickle.loads(bytes_in)
+
+train_x, train_ids, train_y, test_x, test_ids, test_y = data
+
+
+print('loaded data')
+
+if lesion_type == 'none':
+    print('using all features')
+    cat_features = ['token', 'root', 'user',
+                    'prev_token', 'next_token', 'parseroot_token']
+elif lesion_type == 'neighbors':
+    print('lesioning neighbor features')
+    cat_features = ['token', 'root', 'user']
+    for d in train_x + test_x:
+        keys = [key for key in d]
+        for key in keys:
+            if key in ['prev_token', 'next_token', 'parseroot_token']:
+                d.pop(key)
+            if '_pos:' in key:
+                d.pop(key)
+elif lesion_type == 'external':
+    print('lesioning external word features')
+    cat_features = ['token', 'root', 'user',
+                    'prev_token', 'next_token', 'parseroot_token']
+    remove = ['frequency', 'levenshtein', 'leven_frac', 'aoa']
+    for d in train_x + test_x:
+        for key in remove:
+            d.pop(key)
+else:
+    print('Unknown lesion type')
+    sys.exit()
+
+
+# put data in scipy sparse matrix
+dv = DictVectorizer()
+train_x_sparse = dv.fit_transform(train_x)
+test_x_sparse = dv.transform(test_x)
+names = dv.feature_names_
+
+print('features:')
+print(names)
+
+print('built lesioned data set')
+
+params = {
+   'application': 'binary',
+   'metric': 'auc',
+   'learning_rate': .05,
+   'num_leaves': 1024,
+   'min_data_in_leaf': 100,
+   'num_boost_round': 750,
+   'cat_smooth': 200,
+   'max_cat_threshold': 64,
+   'feature_fraction': .7,
[…]

### 2018-03-24 run_lesion.py
add rest of lesion sets

@@ -48,6 +48,25 @@ elif lesion_type == 'external':
     for d in train_x + test_x:
         for key in remove:
             d.pop(key)
+elif lesion_type == 'user':
+    cat_features = ['token', 'root',
+                    'prev_token', 'next_token', 'parseroot_token']
+    remove = ['user', 'entropy', 'burst_length', 'mean_burst_duration', 'median_burst_duration']
+    for d in train_x + test_x:
+        for key in remove:
+            d.pop(key)
+elif lesion_type == 'temporal':
+    cat_features = ['token', 'root', 'user',
+                    'prev_token', 'next_token', 'parseroot_token']
+    key_starters = ['token:', 'root:']
+    key_enders = ['encounters', 'time_since_last_encounter', 'time_since_last_label',
+                  'encounters_lab', 'encounters_unlab', 'first_encounter',
+                  'erravg0', 'erravg1', 'erravg2', 'erravg3']
+    for d in train_x + test_x:
+        for s in key_starters:
+            for e in key_enders:
+                d.pop(s+e, None)
+
 else:
     print('Unknown lesion type')
     sys.exit()

### 2018-03-24 doc/naaclhlt2018-latex/paper.tex
add user features

@@
 Just as we did for word tokens, we encoded the user ID as a single-column,
-high-cardinality feature. We also calculated several other user-level features.
+high-cardinality feature.
+We also calculated several other user-level features that related to the ``learning type'' of a user.
+In particular, we encoded features to estimate the long-term motivation and diligence of a user.
+These features could help predict how users interact with old and novel words they encounter.
 
-\hl{ANSELM HERE}
+%Feature 1: Motivation %compute_motivation
+To estimate users' motivation we grouped their exercises into ``bursts.'' Bursts were separated by at least 24 hours.
+We used three concrete features about these bursts, namely the mean and median number of exercises within bursts as well as the total number of bursts of a given user.
+Users that were more motivated had potentially bursts with more exercises.
+Simultaneously, a larger total number of bursts indicated that the burst length estimate was more trustworthy.
 
-Feature 1: Motivation %compute_motivation
+%Feature 2: Consistency %compute_usage_entropy
 
-Feature 2: Consistency %compute_usage_entropy
+To estimate users' diligence we speculated that a very diligent user might be using the app regularly at the same time of day, perhaps following a study schedule, compared to a less diligent user whose times might vary a lot.
+The data set did not provide a variable with the time of day, which would have been an interesting feature on its own.
+Instead, we were able to extract for each exercise the time of day relative to the first time a user had used the app, ranging from 0 to 1 (with 0 indicating the same time, 0.25 indicating a relative shift by 6 hours, etc.).
+We then discretized this variable into 20-minute bins and computed the entropy of the empirical frequency distribution over these bins.
+A lower entropy score indicated less variability in the times of day a user started their exercises.
 

### 2018-03-26 doc/naaclhlt2018-latex/paper.tex
age of acquisition stuff

@@
 We therefore included a feature that encoded the frequency of each word in the language being acquired, calculated from \citet{robert_speer_2017_998161}.
+A number of studies have also shown that age-of-acquisition is another strong predictor of word processing difficulty that is somewhat independent of word frequency \cite{brysbaert2011effects, ferrand2011comparing}.
+We therefore included mean age-of-acquisition (in English) as a feature, derived from published age-of-acquisition norms for 30,000 English words from \cite{Kuperman2012} which covered many of the words present in the dataset.
 Additionally, cognates, or words sharing a common linguistic derivation, are easier to learn than words with dissimilar translations \cite{de2000hard}.
@@
 We found translations using Google Translate\footnote{https://cloud.google.com/translate/} and calculated the Levenshtein distance to reflect the letter-by-letter similarity of the word and its translation \cite{hyyro2001explaining}.
-\hl{DAVID FILL IN HERE WITH AOA}.
-
 \subsubsection{User features}

### 2018-03-26 doc/naaclhlt2018-latex/paper.tex
clarifying

@@
 We therefore included a feature that encoded the frequency of each word in the language being acquired, calculated from \citet{robert_speer_2017_998161}.
-A number of studies have also shown that age-of-acquisition is another strong predictor of word processing difficulty that is somewhat independent of word frequency \cite{brysbaert2011effects, ferrand2011comparing}.
+A number of studies have also shown that age-of-acquisition is another strong predictor of word processing and lexical retrieval difficulty that is somewhat independent of word frequency \cite{brysbaert2011effects, ferrand2011comparing}.
 We therefore included mean age-of-acquisition (in English) as a feature, derived from published age-of-acquisition norms for 30,000 English words from \cite{Kuperman2012} which covered many of the words present in the dataset.

### 2018-03-26 doc/naaclhlt2018-latex/paper.tex
add lesions graph and description (no interpretation of results yet)

@@
 \usepackage{latexsym}
+\usepackage{graphicx}
 \usepackage{soul}
@@
 the exercise. These included the token of the previous word, the next word, and
-the instance word's root in the parse tree, all stored in single columns as with
+the instance word's root in the dependency tree, all stored in single columns as with
 the instance token itself. We also included the part of speech of each of these
-context words as additional features. When there was no previous word, next word, or parse
+context words as additional features. When there was no previous word, next word, or dependency-tree
 root word, a special {\tt None} token or {\tt None} part of speech was used.
@@
 
+\begin{table}[t!]
+\small
+\begin{center}
+\begin{tabular}{|l|llll|}
+  % \hline   & \multicolumn{3}{l}{\bf Value} & \\
+  \hline \bf Parameter & {\tt fr\_en} & {\tt en\_es} & {\tt es\_en} & {\tt all} \\ \hline
+  num\_leaves & 256 & 512 & 512 & 1024 \\
+  learning\_rate & .05 & .05 & .05 & .05 \\
+  min\_data\_in\_leaf & 100 & 100 & 100 & 100 \\
+  num\_boost\_rounds & 750 & 650 & 600 & 750 \\
+  cat\_smooth & 200 & 200 & 200& 200 \\
+  feature\_fraction & .7 & .7 & .7 & .7 \\
+  max\_cat\_threshold & 32 & 32 & 32& 64 \\
+\hline
+\end{tabular}
+\end{center}
+\caption{\label{lightgbm-params} Parameters of final LightGBM models. See
+  LightGBM documentation for more information; all other parameters were left at
+their default values.}
+\end{table}
+
 After featurizing the training data, we trained gradient boosting decision
@@
 
-\begin{table}[t!]
-\small
-\begin{center}
-\begin{tabular}{|l|llll|}
-  % \hline   & \multicolumn{3}{l}{\bf Value} & \\
-  \hline \bf Parameter & {\tt fr\_en} & {\tt en\_es} & {\tt es\_en} & {\tt all} \\ \hline
-  num\_leaves & 256 & 512 & 512 & 1024 \\
-  learning\_rate & .05 & .05 & .05 & .05 \\
-  min\_data\_in\_leaf & 100 & 100 & 100 & 100 \\
-  num\_boost\_rounds & 750 & 650 & 600 & 750 \\
-  cat\_smooth & 200 & 200 & 200& 200 \\
-  feature\_fraction & .7 & .7 & .7 & .7 \\
-  max\_cat\_threshold & 32 & 32 & 32& 64 \\
-
-\hline
-\end{tabular}
-\end{center}
-\caption{\label{lightgbm-params} Parameters of final LightGBM models. See
-  LightGBM documentation for more information; all other parameters were left at
-their default values.}
-\end{table}
-
 \subsection{Performance}
@@
   es\_en}, and $.8540$ on {\tt fr\_en}. We did not attempt to optimize the model's
-F1 score, but the model's F1 score could likely be improved (at the cost of
+F1 score, but the F1 score could likely be improved (at the cost of
 increased log loss) by finding the
@@
 \section{Feature Removal Experiments}
+
+\begin{figure*}[htp]
+% \centering
+\includegraphics[width=\textwidth]{lesions.pdf}
+\caption{Performance on {\tt dev} of models trained on all {\tt train} data,
+  with different groups of lesioned features. See main text for description of
+  lesion groups}
+\label{fig:lesions}
+\end{figure*}
+
+To better understand which features or groups of features were most important to
+our model's predictions, we conducted a set of experiments in which we lesioned
+(i.e., removed) a group of features and re-trained the model on the {\tt train}
+set, evaluating performance on the {\tt dev} set. For simplicity, we ran each of the
+lesioned models on all language data and report the average
+performance. We did not run individual-language models as we did for our primary model.
+
+The results of the lesion experiments are shown in Figure~\ref{fig:lesions}. The
+models are as follows.
+\begin{itemize}
+  \item \emph{none}: all features are included.  
+  \item \emph{user}: all user-level features, including the user ID and other
+    calculated features like entropy and measures of exercise bursts, are
+    removed.
+  \item \emph{userid} and \emph{user other}: only user ID or only the calculated
+    features user features, respectively, are removed.
+  \item \emph{word}: Token and token root IDs; previous, next, and dependency-tree
+    root word IDs; and morphological, part of speech, and dependency tree
+    features are removed.
+  \item \emph{word id} and \emph{word other}: only word IDs or only other
+    word features, respectively, are removed.
+  \item \emph{neighbors}: Both word IDs and other word features are removed, but
+    only for the previous, next, and dependency-tree root words. Information
+    about the present word is maintained.
+  \item \emph{external}: External information about the word, including corpus
+    frequency, Levenshtein distance from translation, and age of acquisition,
+    are removed.
+  \item \emph{temporal}: Temporal information, including number and timing of
+    past encounters with the word and error tracking information, is removed.
+\end{itemize}
 

### 2018-03-26 graph_lesions.r

@@ -1,0 +1,25 @@
+library('ggplot2')
+
+df = read.csv('data/lesions.csv', header=FALSE,
+              col.names=c('lesion', 'AUC'))
+
+df$lesion = c('none', 'user all', 'user id', 'user other', 'word all', 'word ids', 'word other', 'neighbors', 'external', 'temporal')
+
+df$group = c('none', 'user', 'user', 'user', 'word', 'word', 'word', 'word', 'other', 'other')
+
+df$lesion = factor(df$lesion, levels=df$lesion)
+
+none_score = df[df$lesion == 'none', 'AUC']
+
+
+
+ggplot(df, aes(x=lesion, y=AUC, fill=group)) + geom_bar(stat='identity') +
+  geom_hline(yintercept=none_score) +
+  scale_x_discrete(limits = rev(levels(df$lesion))) +
+  scale_fill_discrete(name='lesion type', breaks=c('none', 'user', 'word', 'other')) +
+  xlab('lesioned features') +
+  ylab('AUROC') +
+  coord_flip(ylim=c(.82, .845)) +
+  theme_minimal()
+
+ggsave('doc/naaclhlt2018-latex/lesions.pdf', height=3, width=8)

### 2018-03-26 run_lesion.py
add more lesion types to script

@@ -40,6 +40,46 @@ elif lesion_type == 'neighbors':
                 d.pop(key)
             if '_pos:' in key:
                 d.pop(key)
+
+elif lesion_type == 'word':
+    print('lesioning word feats')
+    cat_features = ['user']
+    remove = ['token', 'root', 'prev_token', 'next_token', 'parseroot_token']
+    for d in train_x + test_x:
+        keys = [key for key in d]
+        for key in keys:
+            if '_pos:' in key:
+                d.pop(key)
+            elif 'morphological_feature' in key:
+                d.pop(key)
+            elif 'dependency_label' in key:
+                d.pop(key)
+            elif 'part_of_speech' in key:
+                d.pop(key)
+            elif key in remove:
+                d.pop(key)
+elif lesion_type == 'word_ids':
+    print('lesioning word ids')
+    cat_features = ['user']
+    remove = ['token', 'root', 'prev_token', 'next_token', 'parseroot_token']
+    for d in train_x + test_x:
+        for key in remove:
+            d.pop(key, None)
+elif lesion_type == 'word_otherfeats':
+    print('lesioning non-id word features')
+    cat_features = ['token', 'root', 'user',
+                    'prev_token', 'next_token', 'parseroot_token']
+    for d in train_x + test_x:
+        keys = [key for key in d]
+        for key in keys:
+            if '_pos:' in key:
+                d.pop(key)
+            elif 'morphological_feature' in key:
+                d.pop(key)
+            elif 'dependency_label' in key:
+                d.pop(key)
+            elif 'part_of_speech' in key:
+                d.pop(key)
 elif lesion_type == 'external':
     print('lesioning external word features')
     cat_features = ['token', 'root', 'user',
@@ -47,15 +87,31 @@ elif lesion_type == 'external':
     remove = ['frequency', 'levenshtein', 'leven_frac', 'aoa']
     for d in train_x + test_x:
         for key in remove:
-            d.pop(key)
+            d.pop(key, None)
 elif lesion_type == 'user':
+    print('lesioning user features')
     cat_features = ['token', 'root',
                     'prev_token', 'next_token', 'parseroot_token']
     remove = ['user', 'entropy', 'burst_length', 'mean_burst_duration', 'median_burst_duration']
     for d in train_x + test_x:
         for key in remove:
             d.pop(key)
+elif lesion_type == 'user_id':
+    print('lesioning user id')
+    cat_features = ['token', 'root',
[…]

### 2018-03-27 doc/naaclhlt2018-latex/paper.tex
add description of lesions

@@
 models are as follows.
-\begin{itemize}
-  \item \emph{none}: all features are included.  
-  \item \emph{user}: all user-level features, including the user ID and other
+\begin{description}
+  \item[none:] all features are included.  
+  \item[user:] all user-level features, including the user ID and other
     calculated features like entropy and measures of exercise bursts, are
     removed.
-  \item \emph{userid} and \emph{user other}: only user ID or only the calculated
+  \item[userid \& user other:] only user ID or only the calculated
     features user features, respectively, are removed.
-  \item \emph{word}: Token and token root IDs; previous, next, and dependency-tree
+  \item[word:] Token and token root IDs; previous, next, and dependency-tree
     root word IDs; and morphological, part of speech, and dependency tree
     features are removed.
-  \item \emph{word id} and \emph{word other}: only word IDs or only other
+  \item[word id \& word other:] only word IDs or only other
     word features, respectively, are removed.
-  \item \emph{neighbors}: Both word IDs and other word features are removed, but
+  \item[neighbors:] Both word IDs and other word features are removed, but
     only for the previous, next, and dependency-tree root words. Information
     about the present word is maintained.
-  \item \emph{external}: External information about the word, including corpus
+  \item[external:] External information about the word, including corpus
     frequency, Levenshtein distance from translation, and age of acquisition,
     are removed.
-  \item \emph{temporal}: Temporal information, including number and timing of
+  \item[temporal:] Temporal information, including number and timing of
     past encounters with the word and error tracking information, is removed.
-\end{itemize}
+\end{description}
+
+Interestingly, we found that for both user-level and word-level features, the
+bulk of the model's predictive power could be achieved using ID's alone,
+represented as high-cardinality categorical features. Removing other word features,
+such as morphological features and part of speech, created only a small
+degradation of performance. In the case of users, removing features such as
+entropy and average exercise burst length led to a tiny increase of performance.
+In the case of both users and words, though, we find that in the absence of ID
+features the other features are helpful and leads to better performance than
+removing all features. We also found that removing all information about
+neighboring words and the dependency-parse root word degraded performance. This
+confirms that word context matters, and likely points to errors in word order,
+subject--verb matching and other grammatical rules.
+
+Our external word features---Levenshtein distance to translation, frequency, and
+age of acquisition---provided a slight boost to model performance, showing the
+benefit of considering from a psychological and linguistic perspective what
+makes a word hard to learn. Adding temporal features about past encounters and
+errors helped the models, but not as much as we expected. While not included in
+the final model, we had also tried augmenting the temporal feature set with more
+features related to massing and spacing of encounters with a word, but found it
+did not improve performance. This is perhaps not surprising given how small the
+benefit of the existing temporal features are in our model.
+
+While not plotted above, we also ran a model lesioning exercise-level features
+including client, session type, format, and exercise duration. This model
+achieved an AUROC of $.787$, far lower than any other lesion. This points to the
+importance of how a question is asked for user performance, reflecting insights
+from psychology such as the difference between recall and recognition memory
+\cite{yonelinas2002nature}.
 

### 2018-03-27 doc/naaclhlt2018-latex/paper.tex
add discussion

@@
 
+\section{Discussion}
+
+When approaching the Duolingo SLAM task, we hoped to leverage psychological
+insights in building our model. We found that in some cases, such as the
+age of acquisition, this was helpful. In general, though, our model gained its
+power not from hand-crafted features but from applying a powerful inference
+technique (gradient boosted trees) to raw input about user and word IDs and
+exercise features.
+
+There are multiple reasons for the limited applicability of psychology to this
+competition. First, computational psychological models, while useful, are not
+well-suited to generating highly accurate predictions from large data sets.
+Because they are designed not for prediction but for explanation, they tend to
+use a small number of input variables and to allow those variables to interact
+in limited ways. In contrast, gradient boosted trees, as well as other
+cutting-edge techniques like deep learning. can extract high-level
+interactions among hundreds of features. Though they are highly opaque, require
+a lot of data, and are not amenable to explanation, these models excel at prediction.
+
+Second, it is possible that our ability to use theories of learning, including
+ideas about massed and spaced practice, was disrupted by the fact that the data
+may have been adaptively created using these very principles
+\cite{settles2016trainable}. If Duolingo adaptively sequenced the spacing of
+trials based on past errors, then the relationship between future errors and
+past spacing may have differed from that found in the psychological literature
+\cite{Cepeda:2006aa}.
+
+Finally, a task in which more true generalization was required might have
+allowed psychologically inspired features to perform more competitively. In this
+task, there is a large amount of labeled training data for every user and for
+most words. This allow simple ID-based features to work. However, for a
+newly-encountered user or word, an ID is useless. Theory-driven features, in
+contrast, can generalize to new settings. Thus, if we were asked to generalize to
+a completely new language such as German, many parts of our model would falter
+but word frequency, age of acquisition, and Levenshtein distance to
+first-language translation would still likely be predictive.
+
 \section{Acknowledgments}

### 2018-03-27 doc/naaclhlt2018-latex/paper.tex
fixed a bug in description of language pairs

@@
 Educational software that aims to teach people new skills, languages, and
-even academic subjects have become increasingly popular.  The wide-spread
+academic subjects have become increasingly popular.  The wide-spread
 deployment of these tools has created interesting opportunities to study
@@
 predict patterns of future translation mistakes in held-out data.  The data set was
-divided into three sets covering students learning to translate English to Spanish (ES$\to$EN),
-Spanish to English (ES$\to$EN), and French to English (FR$\to$EN).
+divided into three sets covering Spanish speakers learning English ({\tt en\_es}),
+English speakers learning Spanish ({\tt es\_en}), and English speakers learning
+French ({\tt fr\_en}).
 This paper reports the approach used by our team which ended in third place 
-for the ES$\to$EN data set, second place for the ES$\to$EN data set, and third place
-for FR$\to$EN.
+for the {\tt en\_es} data set, second place for {\tt es\_en}, and third place
+for {\tt fr\_en}.
 

### 2018-03-28 doc/naaclhlt2018-latex/paper.tex
proofreading

@@
 
-\author{Alexander S. Rich\qquad Pamela Osborn Popp\qquad David J. Halpern\\
+\author{Alexander S. Rich\qquad Pamela J. Osborn Popp\qquad David J. Halpern\\
   \textbf{Anselm Rothe\qquad Todd M. Gureckis} \\
@@
 language learning app, was
-to use log data from thousand of users completing millions of exercises 
+to use log data from thousands of users completing millions of exercises to 
 predict patterns of future translation mistakes in held-out data.  The data set was
@@
 acquisition has long been a topic of interest in the learning and memory literature~\cite[e.g.,]{Atkinson:1972rm,Atkinson:1972rz,Pavlik:2008rm}.
-In fact, much of the classic work on psychology of memory involved aspects of verbal
+In fact, much of the classic work on the psychology of memory involved aspects of verbal
 language learning, which helped to identify a number of important and robust phenomena
@@
 
-Along with the tokens themselves we encoded \hl{each} instance word's part of speech,
+Along with the tokens themselves we encoded each instance word's part of speech,
 morphological features, and dependency edge label. (We noticed that some words
@@
 We found translations using Google Translate\footnote{https://cloud.google.com/translate/} and calculated the Levenshtein distance to reflect the letter-by-letter similarity of the word and its translation \cite{hyyro2001explaining}.
+
 \subsubsection{User features}
@@
 
-To estimate users' diligence we speculated that a very diligent user might be using the app regularly at the same time of day, perhaps following a study schedule, compared to a less diligent user whose times might vary a lot.
+To estimate users' diligence\hl{consistency?} we speculated that a very diligent user might be using the app regularly at the same time of day, perhaps following a study schedule, compared to a less diligent user whose schedule might vary more.
 The data set did not provide a variable with the time of day, which would have been an interesting feature on its own.
@@
 neighboring words and the dependency-parse root word degraded performance. This
-confirms that word context matters, and likely points to errors in word order,
+confirms that word context matters, and suggests that users commonly make errors in word order,
 subject--verb matching and other grammatical rules.
@@
 Because they are designed not for prediction but for explanation, they tend to
-use a small number of input variables and to allow those variables to interact
+use a small number of input variables and allow those variables to interact
 in limited ways. In contrast, gradient boosted trees, as well as other

### 2018-03-28 doc/naaclhlt2018-latex/paper.tex
some updates to intro

@@
 \usepackage{soul}
+\usepackage{color}
 
@@
 \begin{abstract}
-This is the abstract
+Psychological research on learning and memory
+has tended to emphasize small-scale laboratory
+studies.  However, large datasets of people using 
+educational software provide opportunities to explore
+these issues from a new perspective.  In this paper we
+describe our approach to the Duolingo Second Language
+Acquisition Modeling (SLAM) competition which was run in
+early 2018.   After detailing our modeling approach and a number of
+supplementary simulations, we
+reflect on which aspects of our reasonably
+successful model are substantively
+informed by findings and theories from the psychological literature.
 \end{abstract}
@@
 deployment of these tools has created interesting opportunities to study
-the process of learning in extremely large samples of learners in naturalistic
+the process of language acquisition in extremely large samples of learners in naturalistic
 situations. The Duolingo shared task on Second Lanugage Acquisition Modeling (SLAM)
@@
 
-Our lab's interest in this competition comes from our emphasis on studying human learning
-and memory from a psychological perspective.  While the Duolingo SLAM dataset 
-heavily emphasizes language learning (e.g., syntax, grammar, semantics), it also touches on
-basic learning and memory phenomena (e.g., retrieval of the translated
-word from memory).  The study of the psychological processes supporting second-language 
-acquisition has long been a topic of interest in the learning and memory literature~\cite[e.g.,]{Atkinson:1972rm,Atkinson:1972rz,Pavlik:2008rm}.
-In fact, much of the classic work on the psychology of memory involved aspects of verbal
-language learning, which helped to identify a number of important and robust phenomena
-in memory including the retention curve and the advantage for spaced
-over massed practice~\cite{Ruth:1928aa,Rubin:1996aa,Cepeda:2006aa}.
+Learning and memory has been a core focus of psychological science for
+over 100 years. Most of this work has sought to build explanatory theories of human learning
+and memory using relatively small-scale laboratory studies.  Such studies have
+identified a number of important and apparently robust phenomena in memory including
+the nature of the retention curve~\cite{Rubin:1996aa}, the advantage for spaced over massed 
+practice~\cite{Ruth:1928aa,Cepeda:2006aa,Mozer:2009cs}, the testing effect~\cite{Roediger2006te},
+and retrieval-induced forgetting~\cite{Anderson1994rif}.
+The advent of large datasets such as the one provided in the Duolingo SLAM challenge
+may offer a new perspective and approach which may prove complementary
+to laboratory scale science~\cite{Griffiths:2015aa,Goldstone:2016aa}.  First, the much
+larger sample sizes may help to better identify parameters of psychological models.
+Second, datasets covering more naturalistic learning situations may allow us to test 
+the predictive accuracy of psychological theories in a more generalizable fashion~\cite{Yarkoni:2017aa}.
 
-\hl{ Todd: I want to transition to something about how our approach tried to stick to some
-psychological principles but i'm not sure... maybe need to wait until al the later sections
-are fleshed out to lay out what the "concept" is behind our approach}.
+Despite these promising opportunities, it remains unclear how much of current psychological
+theory might be important for tasks such as the Duolingo SLAM challenge.
+As one example, the field of education data mining which has attempted to build
+predictive models of student learning have found excellent results using
+deep neural networks~\cite[so called ``deep knowledge tracing",][]{Piech:2015aa} 
+as opposed to more traditional, and interpretable, models and approaches
+that are rooted in cognitive science~\cite[e.g.,][]{Atkinson:1972rm,Atkinson:1972rz,Corbett1995bkt,Pavlik:2008rm,Tubridy:2018ab}. 
+Recently, Khajah, Lindsey, \& Mozer (2016) compared deep knowledge tracing (DKT) to a more
+standard ``Bayesian knowledge tracing" (BKT) models and showed
+that it was possible to equate the performance of the the BKT model by additional 
+features and parameters that represent core aspects of the psychology of learning
+and memory such as forgetting~\cite{Khajah2016hdnt}.
 
-\hl{One idea is to talk about while the research on learnign and memory is important
-it might be hard to graft those principles into datasets like these because of 
-subject variability, etc... As a result it is unlikely that straightfoward
-cognitive models will work but instead need to combine maching learning methods
-that can create much more complex rules and decisions with features that reflect
-aspects of what makes something easy or hard to learn.}
+Our entry to SLAM borrowed conceptually from the Khajah et al. (2016) approach.
+Specifically we started with a relatively well known and powerful classification algorithm~\cite[gradient
+boosting decision trees, GBDT,][]{ke2017lightgbm}.  This algorithm has some advantages
+over deep learning approach including the fact that the \hl{inferred decision rules are
+interpretable after the model is fit} while still being able to model relatively complex
+decision criteria.  We then created a number of new features for the SLAM
+the dataset covering aspect such as user perseverance, learning processes, 
+contextual factors, and cognate similarity.  After finding a model which provided the best
+held-out performance on the test data set, we conducted a number of ``lesioning" studies where
+we selectively removed features from the model and re-estimated the parameters in order
+to assess the contribution of particular types of features.  We being by describing our 
+overall modeling approach, and then discuss some fo the lessons learned from our analysis.
 
-\cite{Tubridy:2018aa}
-\cite{Tubridy:2018ab}
-\cite{Corbett1995bkt}
-\cite{Roediger2006te,Anderson1994rif,Mozer:2009cs}
 \section{Task Approach}
@@
 We therefore included a feature that encoded the frequency of each word in the language being acquired, calculated from \citet{robert_speer_2017_998161}.
-A number of studies have also shown that age-of-acquisition is another strong predictor of word processing and lexical retrieval difficulty that is somewhat independent of word frequency \cite{brysbaert2011effects, ferrand2011comparing}.
+A number of studies have also shown that age-of-acquisition (i.e., the age a which children typically exhibit this word in their vocabulary) is another strong predictor of word processing and lexical retrieval difficulty that is somewhat independent of word frequency \cite{brysbaert2011effects, ferrand2011comparing}.
 We therefore included mean age-of-acquisition (in English) as a feature, derived from published age-of-acquisition norms for 30,000 English words from \cite{Kuperman2012} which covered many of the words present in the dataset.
@@
 %Feature 1: Motivation %compute_motivation
-To estimate users' motivation we grouped their exercises into ``bursts.'' Bursts were separated by at least 24 hours.
+To estimate users' motivation we grouped their exercises into ``bursts.'' Bursts were separated by at least 24 hours \hl{todd says: did you change my code because bursts were separate by 1 hour in my imementation}.
 We used three concrete features about these bursts, namely the mean and median number of exercises within bursts as well as the total number of bursts of a given user.
@@
 
-To estimate users' diligence\hl{consistency?} we speculated that a very diligent user might be using the app regularly at the same time of day, perhaps following a study schedule, compared to a less diligent user whose schedule might vary more.
+To estimate users' diligence we speculated that a very diligent user might be using the app regularly at the same time of day, perhaps following a study schedule, compared to a less diligent user whose schedule might vary more.
 The data set did not provide a variable with the time of day, which would have been an interesting feature on its own.
@@
 prior experience with that instance. To capture this, we calculated several
-features related to past experience. We encoded the number of times the
-current exercise's exact sentence had been seen before by the user. 
+features related to past experience.
+
+First, we encoded the number of times the
+current exercise's exact sentence had been seen before by the user. This is
+informed by psychological research showing memory and perceptual processing
+improvement for repeated contexts or chunks~\cite[e.g.,][]{Chun:1999gt}
+
 We also encoded a set of features recording past experience with
@@
 instance token and for the instance root word created by lemmatization.
-\cite{Perruchet:1998nw,Chun:1999gt} \hl{chun is good, maybe something other
-that PARSER as it is more a statistcal learning model}
-
 For each token (and root) we tracked user performance through four weighted
@@
 
-After featurizing the training data, we trained gradient boosting decision
-tree (GBDT) models to minimize log loss. GBDT works by
+After featurizing the training data, we trained GBDT models to minimize log loss. GBDT works by
 iteratively building regression trees, each of which seeks to minimize the
@@
 in limited ways. In contrast, gradient boosted trees, as well as other
-cutting-edge techniques like deep learning. can extract high-level
+cutting-edge techniques like deep learning can extract high-level
 interactions among hundreds of features. Though they are highly opaque, require
@@
 
-Finally, a task in which more true generalization was required might have
+Finally, a task in which braoder generalization was required might have
 allowed psychologically inspired features to perform more competitively. In this
 task, there is a large amount of labeled training data for every user and for
-most words. This allow simple ID-based features to work. However, for a
-newly-encountered user or word, an ID is useless. Theory-driven features, in
-contrast, can generalize to new settings. Thus, if we were asked to generalize to
+most words. This allow simple ID-based features to work because the past
+history of a user will likely influence their future performance. However, for a
+newly-encountered user or word, such an ID is useless because there will be
+not relevant parameter estimate. Theory-driven features, in
+contrast, can often generalize to new settings because they capture more
+generic aspects of the learning task.  For example, if we were asked to generalize to
 a completely new language such as German, many parts of our model would falter
 but word frequency, age of acquisition, and Levenshtein distance to
-first-language translation would still likely be predictive.
+first-language translation would still likely prove to be features which
+have high predictive utility.
+
+In sum, we believe that the Duolingo SLAM dataset and challenge provide
+interesting opportunities for cognitive science and psychology.  One particularly
+useful paradigm is to use large-scale, predictive challenges like this one
+to identify the features or variables that are possibly important for learning.
+Then, complementary laboratory scale studies can be conducted which
+establish the causal status of such features through controlled experimentation.
+\hl{Or something more positive to end on?}
 
 \section{Acknowledgments}
-This research was supported by NSF grant DRL-1631436 and
-seed funds from the NYU Dean for Science.
+This research was supported by NSF grant DRL-1631436 and BCS-1255538,
+and the John S. McDonnell Foundation Scholar Award to TMG.  We thank
+Shannon Tubridy and Tal Yarkoni for helpful suggestions in the development
+of this work.
 

### 2018-03-28 doc/naaclhlt2018-latex/paper.tex
typo

@@
 to use log data from thousands of users completing millions of exercises to 
-predict patterns of future translation mistakes in held-out data.  The data set was
+predict patterns of future translation mistakes in held-out data.  The data was
 divided into three sets covering Spanish speakers learning English ({\tt en\_es}),
@@
 to assess the contribution of particular types of features.  We being by describing our 
-overall modeling approach, and then discuss some fo the lessons learned from our analysis.
+overall modeling approach, and then discuss some of the lessons learned from our analysis.
 

### 2018-03-28 doc/naaclhlt2018-latex/paper.tex
more edits from Alex

@@
 
+\hypersetup{draft}
 \newcommand\BibTeX{B{\sc ib}\TeX}
@@
 Acquisition Modeling (SLAM) competition which was run in
-early 2018.   After detailing our modeling approach and a number of
+early 2018.  We used a well-known class of algorithms (gradient boosted decision
+trees), with features partially informed by theories from the psychological
+literature. After detailing our modeling approach and a number of
 supplementary simulations, we
-reflect on which aspects of our reasonably
-successful model are substantively
-informed by findings and theories from the psychological literature.
+reflect on the degree to which psychological theory aided the model, and
+the potential for cognitive science and predictive modeling competitions to gain
+from each other.
 \end{abstract}
@@
 Our entry to SLAM borrowed conceptually from the Khajah et al. (2016) approach.
-Specifically we started with a relatively well known and powerful classification algorithm~\cite[gradient
-boosting decision trees, GBDT,][]{ke2017lightgbm}.  This algorithm has some advantages
-over deep learning approach including the fact that the \hl{inferred decision rules are
-interpretable after the model is fit} while still being able to model relatively complex
-decision criteria.  We then created a number of new features for the SLAM
-the dataset covering aspect such as user perseverance, learning processes, 
+Specifically we used a relatively well known and powerful classification
+algorithm: gradient boosting decision trees~\cite[GBDT][]{ke2017lightgbm}.  Like
+deep learning, this algorithm can extract complex interactions among features,
+but it has some advantages including faster training and easier integration of
+diverse inputs. We then created a number of new features for the SLAM
+dataset covering aspect such as user perseverance, learning processes, 
 contextual factors, and cognate similarity.  After finding a model which provided the best
@@
 external data sources.
-The word frequency effect suggests that uncommon words are harder to process than common words; readers will look longer at low-frequency words and perform worse in word-identification tasks for these than for high-frequency words \cite{rayner1998eye}.
+Past research suggests that uncommon words are harder to process than common words; readers will look longer at low-frequency words and perform worse in word-identification tasks for these than for high-frequency words \cite{rayner1998eye}.
 We therefore included a feature that encoded the frequency of each word in the language being acquired, calculated from \citet{robert_speer_2017_998161}.
@@
 %Feature 1: Motivation %compute_motivation
-To estimate users' motivation we grouped their exercises into ``bursts.'' Bursts were separated by at least 24 hours \hl{todd says: did you change my code because bursts were separate by 1 hour in my imementation}.
-We used three concrete features about these bursts, namely the mean and median number of exercises within bursts as well as the total number of bursts of a given user.
-Users that were more motivated had potentially bursts with more exercises.
-Simultaneously, a larger total number of bursts indicated that the burst length estimate was more trustworthy.
+To estimate users' motivation we grouped their exercises into ``bursts.'' Bursts
+were separated by at least 24 hours \hl{todd says: did you change my code
+  because bursts were separate by 1 hour in my imementation}.  We used three
+concrete features about these bursts, namely the mean and median number of
+exercises within bursts as well as the total number of bursts of a given user.
+We hypothesized that more motivated users may have bursts with more exercises.
+Simultaneously, a larger total number of bursts indicated that the burst length
+estimate was more trustworthy.
 
@@
 error averages. At the user's first encounter of the token, each error term $E$ starts at
-zero. After an encounter with an instance of the token with label $L$, it is
+zero. After an encounter with an instance of the token with label $L$ (0 for
+success, 1 for error), it is
 updated according to the equation
@@
 
-where $\alpha$ determines the speed of error updating. The four feature weighted
+where $\alpha$ determines the speed of error updating. The four weighted
 error terms use $\alpha = \{.3, .1, .03, .01\}$, allowing both short-run and
@@
 a single split to divide those groups \cite{fisher1958grouping}. Thus, as
-alluded to above, high-cardinality features like token and user were encoded as
+alluded to above, high-cardinality features like token and user ID were encoded as
 single columns and handled as categories by LightGBM.
@@
 To tune model hyper-parameters and evaluate the usefulness of features, we first
-trained the models on the ``train'' data set and evaluated them on the ``dev''
+trained the models on the {\tt train} data set and evaluated them on the {\tt dev}
 data set. Once the model structure was finalized, we trained on the combined
-``train'' and ``dev'' data and produced predictions for the ``test'' data. The
+{\tt train} and {\tt dev} data and produced predictions for the {\tt test} data. The
 LightGBM hyperparameters used for each model are listed in Table~\ref{lightgbm-params}.
@@
 increased log loss) by finding the
-rescaling of the ``dev'' predicted probabilities that maximized the F1 score at
-the 0.5 threshold, and applying this rescaling to the ``test'' predicted probabilities.
+rescaling of the {\tt dev} predicted probabilities that maximized the F1 score at
+the 0.5 threshold, and applying this rescaling to the {\tt test} predicted probabilities.
 
@@
 In the case of both users and words, though, we find that in the absence of ID
-features the other features are helpful and leads to better performance than
+features the other features are helpful and lead to better performance than
 removing all features. We also found that removing all information about
@@
 age of acquisition---provided a slight boost to model performance, showing the
-benefit of considering from a psychological and linguistic perspective what
-makes a word hard to learn. Adding temporal features about past encounters and
+benefit of considering what makes a word hard to learn from a psychological and
+linguistic perspective. Adding temporal features about past encounters and
 errors helped the models, but not as much as we expected. While not included in
@@
 
-While not plotted above, we also ran a model lesioning exercise-level features
+Though not plotted above, we also ran a model lesioning exercise-level features
 including client, session type, format, and exercise duration. This model
@@
 When approaching the Duolingo SLAM task, we hoped to leverage psychological
-insights in building our model. We found that in some cases, such as the
-age of acquisition, this was helpful. In general, though, our model gained its
+insights in building our model. We found that in some cases, such as using the
+word's age of acquisition, this was helpful. In general, though, our model gained its
 power not from hand-crafted features but from applying a powerful inference
-technique (gradient boosted trees) to raw input about user and word IDs and
+technique (gradient boosted decision trees) to raw input about user and word IDs and
 exercise features.
@@
 use a small number of input variables and allow those variables to interact
-in limited ways. In contrast, gradient boosted trees, as well as other
+in limited ways. In contrast, gradient boosted decision trees, as well as other
 cutting-edge techniques like deep learning can extract high-level
@@
 In sum, we believe that the Duolingo SLAM dataset and challenge provide
-interesting opportunities for cognitive science and psychology.  One particularly
-useful paradigm is to use large-scale, predictive challenges like this one
-to identify the features or variables that are possibly important for learning.
-Then, complementary laboratory scale studies can be conducted which
+interesting opportunities for cognitive science and psychology. 
+Large-scale, predictive challenges like this one
+might be used to identify features or variables that are important for learning.
+Then, complementary laboratory-scale studies can be conducted which
 establish the causal status of such features through controlled experimentation.
-\hl{Or something more positive to end on?}
+Conversely, insights from controlled experiments can be used to generate
+new features that aid predictive models on naturalistic datasets. This type of
+two-way interaction could lead to long-run improvements in both causal
+explanation and real-world prediction.
 

### 2018-03-28 doc/naaclhlt2018-latex/paper.tex
some edits

@@
 as opposed to more traditional, and interpretable, models and approaches
-that are rooted in cognitive science~\cite[e.g.,][]{Atkinson:1972rm,Atkinson:1972rz,Corbett1995bkt,Pavlik:2008rm,Tubridy:2018ab}. 
+that are rooted in cognitive science~\cite[e.g.,][]{Atkinson:1972rm,Atkinson:1972rz,Corbett1995bkt,Pavlik:2008rm}.
 Recently, Khajah, Lindsey, \& Mozer (2016) compared deep knowledge tracing (DKT) to a more
@@
 but it has some advantages including faster training and easier integration of
-diverse inputs. We then created a number of new features for the SLAM
+diverse inputs. We then created a number of new psychologically-grounded features for the SLAM
 dataset covering aspect such as user perseverance, learning processes, 
@@
 we selectively removed features from the model and re-estimated the parameters in order
-to assess the contribution of particular types of features.  We being by describing our 
+to assess the contribution of particular types of features.  We begin by describing our 
 overall modeling approach, and then discuss some of the lessons learned from our analysis.
@@
 external data sources.
-Past research suggests that uncommon words are harder to process than common words; readers will look longer at low-frequency words and perform worse in word-identification tasks for these than for high-frequency words \cite{rayner1998eye}.
-We therefore included a feature that encoded the frequency of each word in the language being acquired, calculated from \citet{robert_speer_2017_998161}.
-A number of studies have also shown that age-of-acquisition (i.e., the age a which children typically exhibit this word in their vocabulary) is another strong predictor of word processing and lexical retrieval difficulty that is somewhat independent of word frequency \cite{brysbaert2011effects, ferrand2011comparing}.
-We therefore included mean age-of-acquisition (in English) as a feature, derived from published age-of-acquisition norms for 30,000 English words from \cite{Kuperman2012} which covered many of the words present in the dataset.
+Past research has suggested some features of words that determine how hard words are to process, as measured by how long readers look at the word and performance in lexical decision and word identification tasks.
+Two such features that have somewhat independent effects are word frequency (how often the word occurs in natural language) \cite{rayner1998eye} and age-of-acquisition (the age at which children typically exhibit the word in their vocabulary) \cite{brysbaert2011effects, ferrand2011comparing}.
+We therefore included a feature that encoded the frequency of each word in the language being acquired, calculated from \citet{robert_speer_2017_998161} and a feature that encoded mean age-of-acquisition (in English), derived from published age-of-acquisition norms for 30,000 English words \cite{Kuperman2012} which covered many of the words present in the dataset.
 Additionally, cognates, or words sharing a common linguistic derivation, are easier to learn than words with dissimilar translations \cite{de2000hard}.
@@
 We also calculated several other user-level features that related to the ``learning type'' of a user.
-In particular, we encoded features to estimate the long-term motivation and diligence of a user.
+In particular, we encoded features that might be related to constructs such as the motivation and diligence of a user.
 These features could help predict how users interact with old and novel words they encounter.
@@
 %Feature 1: Motivation %compute_motivation
-To estimate users' motivation we grouped their exercises into ``bursts.'' Bursts
-were separated by at least 24 hours \hl{todd says: did you change my code
+As a proxy for motivation, we speculated that more motivated users would complete more exercises every time they decide to use the app.
+To estimate this, we grouped each user's exercises into ``bursts.'' Bursts were separated by at least 24 hours \hl{todd says: did you change my code
 because bursts were separate by 1 hour in my imementation}.  We used three
 concrete features about these bursts, namely the mean and median number of
-exercises within bursts as well as the total number of bursts of a given user.
-We hypothesized that more motivated users may have bursts with more exercises.
-Simultaneously, a larger total number of bursts indicated that the burst length
-estimate was more trustworthy.
+exercises within bursts as well as the total number of bursts of a given user (to give the model a feature related to the uncertainty in the central tendency estimates).
 
 %Feature 2: Consistency %compute_usage_entropy
-
-To estimate users' diligence we speculated that a very diligent user might be using the app regularly at the same time of day, perhaps following a study schedule, compared to a less diligent user whose schedule might vary more.
+As a proxy for diligence, we speculated that a very diligent user might be using the app regularly at the same time of day, perhaps following a study schedule, compared to a less diligent user whose schedule might vary more.
 The data set did not provide a variable with the time of day, which would have been an interesting feature on its own.
@@
 A lower entropy score indicated less variability in the times of day a user started their exercises.
+An alternative interpretation of the entropy of study schedule parameter described above is that it also relates to context-dependent memory, i.e.
+if the user uses the app at the same time every day they are more likely to be in the same physical location when they are using the app \cite{godden1975context}.
 
@@
 informed by psychological research showing memory and perceptual processing
-improvement for repeated contexts or chunks~\cite[e.g.,][]{Chun:1999gt}
+improvement for repeated contexts or chunks~\cite[e.g.,][]{Chun:1999gt}.
 

### 2018-03-28 doc/naaclhlt2018-latex/paper.tex
edits

@@
 theory might be important for tasks such as the Duolingo SLAM challenge.
-As one example, the field of education data mining which has attempted to build
-predictive models of student learning have found excellent results using
-deep neural networks~\cite[so called ``deep knowledge tracing",][]{Piech:2015aa} 
-as opposed to more traditional, and interpretable, models and approaches
+In the field of education data mining, researchers trying to build
+predictive models of student learning have typically relied on traditional, and interpretable, models and approaches
 that are rooted in cognitive science~\cite[e.g.,][]{Atkinson:1972rm,Atkinson:1972rz,Corbett1995bkt,Pavlik:2008rm}.
-Recently, Khajah, Lindsey, \& Mozer (2016) compared deep knowledge tracing (DKT) to a more
+However, a recent paper found that state-of-the-art results could be achieved using deep neural networks with
+little or no cognitive theory built in~\cite[so called ``deep knowledge tracing",][]{Piech:2015aa} 
+In response, a number of papers such as Khajah, Lindsey, \& Mozer (2016) compared deep knowledge tracing (DKT) to more
 standard ``Bayesian knowledge tracing" (BKT) models and showed
@@
 features and parameters that represent core aspects of the psychology of learning
-and memory such as forgetting~\cite{Khajah2016hdnt}.
+and memory such as forgetting and individual abilities~\cite{Khajah2016hdnt}. An ongoing debate remains in this community 
+whether using flexible models with lots of data can improve over more heavily structured, theory-based models \cite{Tang:2016:DNN, xiong2016going, zhang2017incorporating}.
 
-Our entry to SLAM borrowed conceptually from the Khajah et al. (2016) approach. 
+We positioned our entry to SLAM somewhat in between the above mentioned approaches by using 
+psychologically inspired features in a fairly flexible model. 
 Specifically we used a relatively well known and powerful classification
@@
 external data sources.
-Past research has suggested some features of words that determine how hard words are to process, as measured by how long readers look at the word and performance in lexical decision and word identification tasks.
+Past research in psychology has suggested some features of words that determine how hard the word are to process, as measured by how long readers look at the word and performance in lexical decision and word identification tasks.
 Two such features that have somewhat independent effects are word frequency (how often the word occurs in natural language) \cite{rayner1998eye} and age-of-acquisition (the age at which children typically exhibit the word in their vocabulary) \cite{brysbaert2011effects, ferrand2011comparing}.
@@
 We also calculated several other user-level features that related to the ``learning type'' of a user.
-In particular, we encoded features that might be related to constructs such as the motivation and diligence of a user.
+In particular, we encoded features that might be related to psychological constructs such as the motivation and diligence of a user.
 These features could help predict how users interact with old and novel words they encounter.
@@
 As a proxy for motivation, we speculated that more motivated users would complete more exercises every time they decide to use the app.
-To estimate this, we grouped each user's exercises into ``bursts.'' Bursts were separated by at least 24 hours \hl{todd says: did you change my code
-because bursts were separate by 1 hour in my imementation}.  We used three
+To estimate this, we grouped each user's exercises into ``bursts.'' Bursts were separated by at least an hour.
+We used three
 concrete features about these bursts, namely the mean and median number of
@@
 
-After featurizing the training data, we trained GBDT models to minimize log loss. GBDT works by
+After generating all of the features for the training data, we trained GBDT models to minimize log loss. GBDT works by
 iteratively building regression trees, each of which seeks to minimize the
@@
     root word IDs; and morphological, part of speech, and dependency tree
-    features are removed.
+    features are removed. This does not include {\bf external} features listed below.
   \item[word id \& word other:] only word IDs or only other
-    word features, respectively, are removed.
+    word features, respectively, are removed. {\bf word other} does not include {\bf external} features listed below.
   \item[neighbors:] Both word IDs and other word features are removed, but
@@
 insights in building our model. We found that in some cases, such as using the
-word's age of acquisition, this was helpful. In general, though, our model gained its
+word's age-of-acquisition, this was helpful. In general, though, our model gained its
 power not from hand-crafted features but from applying a powerful inference
@@
 
-Finally, a task in which braoder generalization was required might have
+Finally, a task in which broader generalization was required might have
 allowed psychologically inspired features to perform more competitively. In this
@@
 history of a user will likely influence their future performance. However, for a
-newly-encountered user or word, such an ID is useless because there will be
-not relevant parameter estimate. Theory-driven features, in
+newly-encountered user or word, such an ID is useless because there will not be
+enough data to estimate the relevant parameter. Theory-driven features, in
 contrast, can often generalize to new settings because they capture more

### 2018-03-29 doc/naaclhlt2018-latex/paper.tex
edits from Anselm

@@
 
-We positioned our entry to SLAM somewhat in between the above mentioned approaches by using 
-psychologically inspired features in a fairly flexible model. 
-Specifically we used a relatively well known and powerful classification
-algorithm: gradient boosting decision trees~\cite[GBDT][]{ke2017lightgbm}.  Like
-deep learning, this algorithm can extract complex interactions among features,
+For our approach to the SLAM competition, we decided to use a generic and fairly flexible model structure that we provided with hand-coded, psychologically inspired features.
+We therefore positioned our entry to SLAM somewhat in between the approaches mentioned above.
+Specifically, we used gradient boosting decision trees~\cite[GBDT][]{ke2017lightgbm} for the model structure, which is a powerful classification algorithm that is known to perform well across various kinds of data sets.
+Like
+deep learning, GBDT can extract complex interactions among features,
 but it has some advantages including faster training and easier integration of
-diverse inputs. We then created a number of new psychologically-grounded features for the SLAM
+diverse inputs.
+
+We then created a number of new psychologically-grounded features for the SLAM
 dataset covering aspect such as user perseverance, learning processes, 
-contextual factors, and cognate similarity.  After finding a model which provided the best
+contextual factors, and cognate similarity.  After finding a model that provided the best
 held-out performance on the test data set, we conducted a number of ``lesioning" studies where
@@
 
-We approached the task as a binary classification problem over instances (i.e.,
-single words within an exercise). Our solution can be divided into two
-components---constructing a set of features that is highly informative about
-whether the user will answer an instance correctly, and designing a model that
+We approached the task as a binary classification problem over instances. Each instance was
+a single word within a sentence of a translation exercise and the classification problem was to predict whether a user would translate the word correctly or not. Our approach can be divided into two
+components---constructing a set of features that is informative about
+whether a user will answer an instance correctly, and designing a model that
 can achieve high performance using this feature set.
@@
 Along with the tokens themselves we encoded each instance word's part of speech,
-morphological features, and dependency edge label. (We noticed that some words
+morphological features, and dependency edge label. We noticed that some words
 in the original dataset were paired with the wrong morphological features,
 particularly near where punctuation had been removed from the sentence. To fix
-this, we re-processed the data using Google SyntaxNet\footnote{https://github.com/ljm625/syntaxnet-rest-api}.)
+this, we re-processed the data using Google SyntaxNet\footnote{https://github.com/ljm625/syntaxnet-rest-api}.
 
@@
 external data sources.
-Past research in psychology has suggested some features of words that determine how hard the word are to process, as measured by how long readers look at the word and performance in lexical decision and word identification tasks.
-Two such features that have somewhat independent effects are word frequency (how often the word occurs in natural language) \cite{rayner1998eye} and age-of-acquisition (the age at which children typically exhibit the word in their vocabulary) \cite{brysbaert2011effects, ferrand2011comparing}.
-We therefore included a feature that encoded the frequency of each word in the language being acquired, calculated from \citet{robert_speer_2017_998161} and a feature that encoded mean age-of-acquisition (in English), derived from published age-of-acquisition norms for 30,000 English words \cite{Kuperman2012} which covered many of the words present in the dataset.
-Additionally, cognates, or words sharing a common linguistic derivation, are easier to learn than words with dissimilar translations \cite{de2000hard}.
+Research in psychology has suggested certain word features that play a role in how difficult a word is to process, as measured by how long readers look at the word as well as people's performance in lexical-decision and word-identification tasks.
+Two such features that have somewhat independent effects are word frequency \cite[i.e., how often does the word occur in natural language;][]{rayner1998eye} and age-of-acquisition \cite[i.e., the age at which children typically exhibit the word in their vocabulary;][]{brysbaert2011effects, ferrand2011comparing}.
+We therefore included a feature that encoded the frequency of each word in the language being acquired, calculated from \citet{robert_speer_2017_998161}, and a feature that encoded the mean age-of-acquisition (of the English word, in English native speakers), derived from published age-of-acquisition norms for 30,000 words \cite{Kuperman2012}, which covered many of the words present in the dataset.
+Additionally, words sharing a common linguistic derivation (also called ``cognates''; e.g., ``secretary'' in English and ``secretario'' in Spanish), are easier to learn than words with dissimilar translations \cite{de2000hard}.
 As an approximate measure of linguistic similarity, we used the Levenshtein edit distance between the word tokens and their translations scaled by the length of the longer word.
@@
 A lower entropy score indicated less variability in the times of day a user started their exercises.
-An alternative interpretation of the entropy of study schedule parameter described above is that it also relates to context-dependent memory, i.e.
-if the user uses the app at the same time every day they are more likely to be in the same physical location when they are using the app \cite{godden1975context}.
+
+The entropy score might also give an indication for context effects on users' memory.
+A user practicing exercises more regularly is more likely to be in the same physical location when using the app, which might result in better memory of previously studied words \cite{godden1975context}.
 
@@
 success, 1 for error), it is
-updated according to the equation
+updated according to the equation:
 
@@
 trained the models on the {\tt train} data set and evaluated them on the {\tt dev}
-data set. Once the model structure was finalized, we trained on the combined
+data set.
+\hl{Anselm: if space permits, mention what the train, dev, and test data sets look like?
+e.g.:  80\% of each user's exercises are in train, 10\% in each of dev and test.} Once the model structure was finalized, we trained on the combined
 {\tt train} and {\tt dev} data and produced predictions for the {\tt test} data. The
@@
 
+\hl{Anselm: mention AUROC of baseline model for comparison?
+shall the reader be impressed by our performance.
+perhaps also mention the winner's score (the deep learning team) for commparison, to show that our score was not that much worse?
+`we were 0.01 behind the deep learning model' sounds cooler to me than `we were third'}
 The AUROC of our final predictions was $.8585$ on {\tt en\_es}, $.8350$ on {\tt
@@
 performance. We did not run individual-language models as we did for our primary model.
-
 The results of the lesion experiments are shown in Figure~\ref{fig:lesions}. The
@@
 \begin{description}
-  \item[none:] all features are included.  
-  \item[user:] all user-level features, including the user ID and other
+  \item[none:] All features are included.  
+  \item[user:] All user-level features, including the user ID and other
     calculated features like entropy and measures of exercise bursts, are
     removed.
-  \item[userid \& user other:] only user ID or only the calculated
+  \item[userid \& user other:] Only user ID or only the calculated
     features user features, respectively, are removed.
@@
     features are removed. This does not include {\bf external} features listed below.
-  \item[word id \& word other:] only word IDs or only other
+  \item[word id \& word other:] Only word IDs or only other
     word features, respectively, are removed. {\bf word other} does not include {\bf external} features listed below.
@@
 \cite{yonelinas2002nature}. 
+\hl{Anselm: I don't understand the last sentence here.}
 
@@
 When approaching the Duolingo SLAM task, we hoped to leverage psychological
-insights in building our model. We found that in some cases, such as using the
+insights in building our model. We found that in some cases, such as when using the
 word's age-of-acquisition, this was helpful. In general, though, our model gained its
 power not from hand-crafted features but from applying a powerful inference
-technique (gradient boosted decision trees) to raw input about user and word IDs and
+technique (gradient boosted decision trees) to raw input about user IDs, word IDs, and
 exercise features.
@@
 There are multiple reasons for the limited applicability of psychology to this
-competition. First, computational psychological models, while useful, are not
-well-suited to generating highly accurate predictions from large data sets.
+competition. First, computational psychological models are typically designed based 
+on small laboratory data sets, which might limit their suitability for generating highly
+accurate predictions in big data settings.
 Because they are designed not for prediction but for explanation, they tend to
@@
 cutting-edge techniques like deep learning can extract high-level
-interactions among hundreds of features. Though they are highly opaque, require
+interactions among hundreds of features. While they are highly opaque, require
 a lot of data, and are not amenable to explanation, these models excel at prediction.
@@
 trials based on past errors, then the relationship between future errors and
-past spacing may have differed from that found in the psychological literature
+past spacing may have substantially differed from that found in the psychological literature
 \cite{Cepeda:2006aa}.
 
-Finally, a task in which broader generalization was required might have
-allowed psychologically inspired features to perform more competitively. In this
+Finally, if the task had required broader generalization, psychologically inspired 
+features might have performed more competitively. In the SLAM
 task, there is a large amount of labeled training data for every user and for
-most words. This allow simple ID-based features to work because the past
-history of a user will likely influence their future performance. However, for a
-newly-encountered user or word, such an ID is useless because there will not be
-enough data to estimate the relevant parameter. Theory-driven features, in
-contrast, can often generalize to new settings because they capture more
-generic aspects of the learning task.  For example, if we were asked to generalize to
+most words. This allows simple ID-based features to work because the past
+history of a user will likely influence their future performance. However, with
+ ID-based features there is no way to generalize to newly-encountered users or 
+ words, which have an ID that was not in the training set. The learned 
+ID-based knowledge is useless here because there is no way to generalize from one 
+unique ID to another. Theory-driven features, in
+contrast, can often generalize to new settings because they capture aspects that 
+are shared across (subsets of) users, words, or situations of the learning task.  For example, if we were asked to generalize to
 a completely new language such as German, many parts of our model would falter
@@
 have high predictive utility.
+\hl{I like that sentence! Bam take that, deep learning}
 
@@
 new features that aid predictive models on naturalistic datasets. This type of
-two-way interaction could lead to long-run improvements in both causal
+two-way interaction could lead to long-run improvements in both scientific
 explanation and real-world prediction.

### 2018-03-29 doc/naaclhlt2018-latex/paper.tex
ok

@@
 deployment of these tools has created interesting opportunities to study
-the process of language acquisition in extremely large samples of learners in naturalistic
-situations. The Duolingo shared task on Second Lanugage Acquisition Modeling (SLAM)
+the process of learning in large samples. The Duolingo shared task on Second Lanugage Acquisition Modeling (SLAM)
 was a competitive modeling challenge run in early 2018~\cite{slam18}.
@@
 However, a recent paper found that state-of-the-art results could be achieved using deep neural networks with
-little or no cognitive theory built in~\cite[so called ``deep knowledge tracing",][]{Piech:2015aa} 
-In response, a number of papers such as Khajah, Lindsey, \& Mozer (2016) compared deep knowledge tracing (DKT) to more
+little or no cognitive theory built in~\cite[so called ``deep knowledge tracing",][]{Piech:2015aa}.
+Khajah, Lindsey, \& Mozer (2016) compared deep knowledge tracing (DKT) to more
 standard ``Bayesian knowledge tracing" (BKT) models and showed
@@
 We therefore positioned our entry to SLAM somewhat in between the approaches mentioned above.
-Specifically, we used gradient boosting decision trees~\cite[GBDT][]{ke2017lightgbm} for the model structure, which is a powerful classification algorithm that is known to perform well across various kinds of data sets.
+Specifically, we used gradient boosting decision trees~\cite[GBDT,][]{ke2017lightgbm} for the model structure, which is a powerful classification algorithm that is known to perform well across various kinds of data sets.
 Like
@@
 We then created a number of new psychologically-grounded features for the SLAM
-dataset covering aspect such as user perseverance, learning processes, 
+dataset covering aspects such as user perseverance, learning processes, 
 contextual factors, and cognate similarity.  After finding a model that provided the best
@@
 informed by psychological research showing memory and perceptual processing
-improvement for repeated contexts or chunks~\cite[e.g.,][]{Chun:1999gt}.
+improvements for repeated contexts or ``chunks"~\cite[e.g.,][]{Chun:1999gt}.
 
@@
 trained the models on the {\tt train} data set and evaluated them on the {\tt dev}
-data set.
-\hl{Anselm: if space permits, mention what the train, dev, and test data sets look like?
-e.g.:  80\% of each user's exercises are in train, 10\% in each of dev and test.} Once the model structure was finalized, we trained on the combined
+data set. Details of the training sets and the actual files are provided on the
+Harvard Dataverse\footnote{https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/8SWHNO}.
+Once the model structure was finalized, we trained on the combined
 {\tt train} and {\tt dev} data and produced predictions for the {\tt test} data. The
@@
 
-\hl{Anselm: mention AUROC of baseline model for comparison?
-shall the reader be impressed by our performance.
-perhaps also mention the winner's score (the deep learning team) for commparison, to show that our score was not that much worse?
-`we were 0.01 behind the deep learning model' sounds cooler to me than `we were third'}
 The AUROC of our final predictions was $.8585$ on {\tt en\_es}, $.8350$ on {\tt
-  es\_en}, and $.8540$ on {\tt fr\_en}. We did not attempt to optimize the model's
+  es\_en}, and $.8540$ on {\tt fr\_en}. For reference this placed us within .01 of the winning
+  entry for each problem ($.8613$ on {\tt en\_es}, $.8383$ on {\tt
+  es\_en}, and $.8570$ on {\tt fr\_en}).  Also note that the Duolingo-provided baseline
+  model (L2-regularized regression trained with stochastic gradient descent weighted
+  by frequency) obtains $.7737$ on {\tt en\_es}, $.7456$ on {\tt
+  es\_en}, and $.7707$ on {\tt fr\_en}.
+  We did not attempt to optimize the model's
 F1 score, but the F1 score could likely be improved (at the cost of
@@
 achieved an AUROC of $.787$, far lower than any other lesion. This points to the
-importance of how a question is asked for user performance, reflecting insights
+importance of how memory is assess on performance, reflecting insights
 from psychology such as the difference between recall and recognition memory
 \cite{yonelinas2002nature}.
-\hl{Anselm: I don't understand the last sentence here.}
 
@@
 have high predictive utility.
-\hl{I like that sentence! Bam take that, deep learning}
 
@@
 Conversely, insights from controlled experiments can be used to generate
-new features that aid predictive models on naturalistic datasets. This type of
+new features that aid predictive models on naturalistic datasets~\cite{Griffiths:2015aa,Goldstone:2016aa}. This type of
 two-way interaction could lead to long-run improvements in both scientific

### 2018-03-29 doc/naaclhlt2018-latex/paper.tex
never ends

@@
 standard ``Bayesian knowledge tracing" (BKT) models and showed
-that it was possible to equate the performance of the the BKT model by additional 
+that it was possible to equate the performance of the BKT model by additional 
 features and parameters that represent core aspects of the psychology of learning
@@
 trained the models on the {\tt train} data set and evaluated them on the {\tt dev}
-data set. Details of the training sets and the actual files are provided on the
+data set. Details of the datasets and the actual files are provided on the
 Harvard Dataverse\footnote{https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/8SWHNO}.
@@
 achieved an AUROC of $.787$, far lower than any other lesion. This points to the
-importance of how memory is assess on performance, reflecting insights
-from psychology such as the difference between recall and recognition memory
-\cite{yonelinas2002nature}.
+fact that \emph{how} memory is assessed often includes observed performance 
+\cite[e.g., the large literature in psychology on the difference between recall and recognition memory,][]{yonelinas2002nature}.
 

### 2018-04-10 translate_frequency.py
created translate_frequency.py

@@ -1,0 +1,202 @@
+# -*- coding: utf-8 -*-
+"""
+word features:
+    
+    This script processes the data files to create features about translations 
+    and frequencies.
+    Specifically, word frequency, levenshtein distance (cognate), age of acquisition
+    
+    Saves features as .txt file for specified language
+    There is code to do this for the root words (e.g. "is" becomes "be") and code 
+    to do this for all words ("be" and "is" as two separate words).
+    
+    March 2018
+"""
+
+# imports
+# do pip installs to get libraries: wordfreq, googletrans, editdistance, unidecode
+
+from wordfreq import word_frequency
+from googletrans import Translator
+import editdistance
+from unidecode import unidecode
+import time
+import pandas as pd
+import numpy as np
+
+
+translator = Translator()
+
+# choose language to translate
+lang='en_es'
+fn = 'data/data_{0}/{0}.slam.20171218.train.new'.format(lang)
+(src, dest)=lang.split('_')
+print("Translating %s to %s" %(src,dest))
+
+aoa_xl = pd.ExcelFile("data/13428_2013_348_MOESM1_ESM.xlsx")
+aoa_df = aoa_xl.parse('Sheet1')
+
+# FOR EVERY ROOT WORD (e.g, "be" instead of "is")
+# Make csv of translation and calculate a simple metric of how much of a "cognate" it is
+
+#output filename
+fnout = "{}_{}_rootwordfeats.txt".format(src,dest)
+print("Will save translations and features to: ",fnout)
+
+
+trans_cache = {}
+
+newlines=[]
+
+with open(fn, 'r') as f:
+    for line in f:
+        if len(line)>1 and line[0]!='#':
+            # process and translate
+            pieces = line.split()
+            idx = pieces[0]
+            word = pieces[1]
+            root_token = pieces[2]
+            part_of_speech = pieces[3]
+            morph_feats = pieces[4]
+            dependency_label = pieces[5]
+            dependency_edge_head = pieces[6]
+
+            clean_word = root_token.lower() # Only using root token here rather than word
+            if '\'' not in clean_word and clean_word not in trans_cache:
+                # Wait so as not to annoy the API (otherwise will give JSON error)
+                time.sleep(0.2)
+
+                # Translate so we know the word in the language the user already knows
+                translation_obj = translator.translate(clean_word,src=src,dest=dest)
+                translation = translation_obj.text
+
+                # Clean up translation: 
+                # remove accents, make lowercase, and remove any non-informative first words
+                clean_trans = unidecode(translation)
[…]

### 2018-04-12 doc/naaclhlt2018-latex/paper.tex
addressed reviews, comments welcome

@@
 
-\hypersetup{draft}
+% \hypersetup{draft}
 \newcommand\BibTeX{B{\sc ib}\TeX}
@@
 French ({\tt fr\_en}).
-This paper reports the approach used by our team which ended in third place 
+This paper reports the approach used by our team, which finished in third place 
 for the {\tt en\_es} data set, second place for {\tt es\_en}, and third place
@@
 are unavailable, so predictions must be made using a mix of labeled and
-unlabeled past encounters. To generate training-set features that are 
+unlabeled past encounters. In particular, for a user's test set with $n$
+exercises, each exercise will have between zero and $n-1$ preceding unlabeled exercises.
+
+To generate training-set features that are 
 comparable to test-set features, we selectively ignored some labels when encoding temporal features on
 the training set. Specifically, for each user we first calculated the number of
-exercises $n$ in the true test set. Then, when encoding the features for each
-instance, we selected a random integer $r$ in the range $[1,n]$, and ignored labels
-in the prior $r$ exercises. That is, we encode features for the current instance
-as though other instances in those prior exercises were unlabeled, and ignore
+exercises $n$ in their true test set\footnote{If the size of the test set were
+  not available, it could be estimated based on the fact that it is
+  approximately 5\% of each participant's data.}. Then, when encoding the features for each
+training instance, we selected a random integer $r$ in the range $[0,n-1]$, and
+ignored the labels in the prior $r$ exercises. That is, we encoded features for the instance
+as though other instances in those prior exercises were unlabeled, and ignored
 updates to the error averages from those exercises. The result of this process
-is that each instance in the training set is encoded as though it were between
+is that each instance in the training set was encoded as though it were between
 one and $n$ exercises into the test set.
@@
   \hline \bf Parameter & {\tt fr\_en} & {\tt en\_es} & {\tt es\_en} & {\tt all} \\ \hline
-  num\_leaves & 256 & 512 & 512 & 1024 \\
-  learning\_rate & .05 & .05 & .05 & .05 \\
-  min\_data\_in\_leaf & 100 & 100 & 100 & 100 \\
-  num\_boost\_rounds & 750 & 650 & 600 & 750 \\
-  cat\_smooth & 200 & 200 & 200& 200 \\
-  feature\_fraction & .7 & .7 & .7 & .7 \\
-  max\_cat\_threshold & 32 & 32 & 32& 64 \\
+  {\tt num\_leaves} & 256 & 512 & 512 & 1024 \\
+  {\tt learning\_rate} & .05 & .05 & .05 & .05 \\
+  {\tt min\_data\_in\_leaf} & 100 & 100 & 100 & 100 \\
+  {\tt num\_boost\_rounds} & 750 & 650 & 600 & 750 \\
+  {\tt cat\_smooth} & 200 & 200 & 200& 200 \\
+  {\tt feature\_fraction} & .7 & .7 & .7 & .7 \\
+  {\tt max\_cat\_threshold} & 32 & 32 & 32& 64 \\
 \hline
@@
 training, we averaged the predictions of each single-language model with that of
-the all-language model to form our final predictions.
+the all-language model to form our final predictions. Informal experimentation
+showed that model averaging provided a modest performance boost, and that
+weighted averages did not clearly outperform a simple average.
 
@@
   es\_en}, and $.7707$ on {\tt fr\_en}.
-  We did not attempt to optimize the model's
-F1 score, but the F1 score could likely be improved (at the cost of
-increased log loss) by finding the
-rescaling of the {\tt dev} predicted probabilities that maximized the F1 score at
-the 0.5 threshold, and applying this rescaling to the {\tt test} predicted probabilities.
+We did not attempt to optimize F1 score, the competition's secondary evaluation metric.
+%   We did not attempt to optimize the model's
+% F1 score., but the F1 score could likely be improved (at the cost of
+% increased log loss) by finding the
+% rescaling of the {\tt dev} predicted probabilities that maximized the F1 score at
+% the 0.5 threshold, and applying this rescaling to the {\tt test} predicted probabilities.
 
@@
   \item[none:] All features are included.  
-  \item[user:] All user-level features, including the user ID and other
+  \item[user all:] All user-level features, including the user ID and other
     calculated features like entropy and measures of exercise bursts, are
     removed.
-  \item[userid \& user other:] Only user ID or only the calculated
-    features user features, respectively, are removed.
-  \item[word:] Token and token root IDs; previous, next, and dependency-tree
+  \item[user id \& user other:] Only user ID or only the calculated
+    user features, respectively, are removed.
+  \item[word all:] Token and token root IDs; previous, next, and dependency-tree
     root word IDs; and morphological, part of speech, and dependency tree
@@
 achieved an AUROC of $.787$, far lower than any other lesion. This points to the
-fact that \emph{how} memory is assessed often includes observed performance 
+fact that the manner in which memory is assessed often affects observed performance 
 \cite[e.g., the large literature in psychology on the difference between recall and recognition memory,][]{yonelinas2002nature}.
